---
title: "Lab-02: Getting Started with R"
author: Danni CHEN
date: Assignment due by 11:59 PM on Sunday, 9/8/2019
output:
  html_document:
    toc: FALSE
    toc_depth: 2
    toc_float: true
    theme: simplex
    fig_caption: true
    number_sections: true
---
`

# R Syntax
Why do the following code chunks not work? In each case, correct the code and briefly explain the cause of the error.

Note: you can run each code chunk by itself by clicking the "Run" button at the top right of the code chunk, or using the shortcut Ctrl+Shift+Enter. 

## Variables
Correct any syntax errors in the following code chunk.
```{r}
my_variable <- 10
print(my_variable)
```
ANSWER: The spelling of "my_variable" in the print code is wrong.


## Relational operations
Why doesn't R think that `0.3 + 0.6` equals `0.9`?
```{r}
0.3 + 0.6 == 0.9

0.3+0.6-0.9
```
ANSWER: In R, the result of 0.3 plus 0.6 is not exactly equal to 0.9, but nearly equal to 0.9.



In the following code chunk, write a statement to test if two numbers are nearly equal. Use it to relate `0.3 + 0.6` to `0.9`. Note: a good solution should report either `TRUE` or `FALSE`.
```{r}
all.equal(0.3+0.6, 0.9)
```

## Conditional statement
Correct any syntax errors in the following code chunk.
```{r}

if (1 > FALSE) {
  print("1 > FALSE")
} else {
  print("1 <= FALSE")
}
```

## Loop
Correct any syntax errors in the following code chunk.
```{r}
for (i in list("a", TRUE, 0)) (
  print(i)
)
```


# Subsetting
Below each comment in the following code chunk, add code to perform the requested [subsetting](https://adv-r.hadley.nz/subsetting.html) operations.
```{r}
v <- c(3, 1, 8, 45, 2, 4, 0)

# Use positive integers to subset v to the 2nd and 5th elements
  v[c(2,5)]

# Use negative integers to exclude the 2nd and 5th elements of v
  v[-c(2,5)]

# Use a logical vector to select the first and last elements of v
  v[c(TRUE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE)]

# Use the following expression to subset v to elements that are greater than 2
  v[v > 2]


# Use the head() function to display the first 6 elements of v
  head(v, 6)

# Use the head() function to display the first 3 elements of v
  head(v, 3)

```


# Variable Operations
Add to or modify the code in each code chunk to implement the step described above each chunk.

1. Create a vector `y` with the following values: 5, -2, 31, 4, 32, 7, -9.
```{r}
y <- c(5, -2, 31, 4, 32, 7, -9)
```

2. Calculate the average value of `y`, and assign it to a new variable `y_mean`.
```{r}
y_mean <- mean(y)

y_mean
```

3. Create a new vector `y_error` which gives the difference between each value of `y` and `y_mean`. 
```{r}
y_error <- y - y_mean

y_error
```

4. Using the `length()` function, construct a single expression that confirms two things: (a) `y_error` and `y` both contain the same number of elements, and (b) the mean of `y_error` is equal to zero (up to machine precision). In other words, evaluating this expression should generate a value of `TRUE` if both conditions (a) and (b) are true, but a value of `FALSE` if either condition is false.
```{r}
# (a)
      length(y_error)==length(y)

# (b)
      mean(y_error)==0

# (a) and (b)
      length(y_error)==length(y) & mean(y_error)==0
```

5. Create a new vector `y_abs_error` which gives the absolute value of element of `y_error`. Calculate and print the mean of the absolute error variable.
```{r}
y_abs_error <- abs(y_error)
mean(y_abs_error)
  
```

# Functions
Create a function called `to_zero` that takes a numeric value `x` and subtracts 1 if the input value is greater than 0, adds 1 if the input is less than zero, and returns zero if the input is equal to zero. Follow the comments below.
```{r}
# Define the function to use a default value of x = 2
to_zero <- function(x) {
  if (x>0){
    x=x-1
  }
  else if (x<0){
    x=x+1
  }
  x
}
# Evaluate the function with no argument. Also evaluate the function at the values 1, 0, and -1.
to_zero(20)
to_zero(1)
to_zero(0)
to_zero(-1)
```


# Data Frames
The following exercises involve reading text data into a data frame. You may wish to refer to the lecture handout or to the [Data import](https://r4ds.had.co.nz/data-import.html) chapter of the textbook.

## Read a delimited (csv) text file into a data frame
Use the `read_csv()` function from the `tidyverse` package to load the CSV file `20160819_OlympicSportsByCountries.csv` available at the web address below. For R to recognize the `read_csv` command, you will need to add a line of code to load the `tidyverse` package, e.g. by using the `require` command. 

**HINT:** After the code chunk below has been correctly modified, the data frame `trends_csv` should have three variables (Country, Search Interest, and Sport) and 1,408 rows. You will need to modify the arguments of the `read_csv` command below to read in the data file without error. You may want to open the raw data file in a text editor to gain insight into what may be causing the error. Enter `?read_csv` in the R Console to learn more about the `read_csv` command. 

Original code:
```{r}
url <- "https://raw.githubusercontent.com/googletrends/data/master/20160819_OlympicSportsByCountries.csv"
require(tidyverse)
trends_csv <- read_csv(url, skip=2, col_names = TRUE, col_types=cols(
  Country = col_character(),
  `Search Interest` = col_double(),
  Sport = col_character()), n_max = 1408
)
all.equal(names(trends_csv), c("Country", "Search Interest", "Sport"))
nrow(trends_csv)==1408
```


## Preview data in data frame
Print the first three observations (i.e. first three rows) in the data frame `trends_csv` created above. Do this using the `head` function, and also do it using index notation as described in the section on Selecting Observations [here](https://www.statmethods.net/management/subset.html). 
```{r}
head(trends_csv, 3)
trends_csv[1:3,]
```

## Subsetting a data frame
Using the data frame `trends_csv` created above, select and list all observations corresponding to the country of China. Which sport is most searched for in China?
```{r}
subset(trends_csv, Country == "China")

```
ANSWER: Swimming is most searched for in China.

## Going further
**Advanced (and optional):** Write a function called `top_sport()` that takes the name of a country as a single (string) argument and prints the name of the most searched for sport in that country based on data in the data frame `trends_csv` created above. E.g., `top_sport("United States")` should print the name of the most searched for sport in the U.S. If you figure this out, feel free to email the professor your proposed solution!
```{r}
top_sport <- function(x) {
  This_country = subset(trends_csv, Country == x)
  top=This_country[which.max(This_country$`Search Interest`),3]
  top
}
top_sport("United States")
top_sport("China")
top_sport("Russia")
```


